---
title: "HIV_Prediction"
author: "Amanuel Belay"
date: "2024-09-01"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(skimr)
library(tidymodels)
library(GGally)
library(usemodels)
library(vip)
library(themis)
```

```{r}
hiv_bio <- read_csv("data/ephia2017adultbio.csv")
hiv_int <- read_csv ("data/ephia2017adultind.csv")
```

```{r}
skim(hiv_bio)
```

# merge the two data sets

```{r}
hiv_status <- hiv_bio %>%
  select(personid, bt_status, hiv1statusfinalsurvey) 
```

```{r}
hiv_full <- merge(hiv_int, hiv_status, by = "personid") %>%
  mutate(hiv_result = case_when(hiv1statusfinalsurvey == 1 ~ "positive",
                                hiv1statusfinalsurvey == 2 ~ "negative"),
         gender = case_when(gender == 1 ~ "male",
                            gender == 2 ~ "female"),
         region = case_when(region == 1 ~ "Tigray",
                            region == 2 ~ "Afar",
                            region == 3 ~ "Amhara",
                            region == 4 ~ "Oromia",
                            region == 5 ~ "Somali",
                            region == 6 ~ "Benishangul Gumuz",
                            region == 7 ~ "SNNPR",
                            region == 8 ~ "Gambella",
                            region == 9 ~ "Harari",
                            region == 10 ~ "Addis Ababa",
                            region == 11 ~ "Dire Dawa")) %>%
  select(-hiv1statusfinalsurvey) %>% 
  filter(!is.na(hiv_result))
```

# Explor the data (EDA)
```{r}
hiv_full %>% count(hiv_result)
```

```{r}
hiv_full %>% filter(!is.na(age)) %>%
  ggplot(aes(x = hiv_result, y = age, fill = hiv_result)) +
  geom_boxplot()
```

```{r}
hiv_full %>% count(gender)

hiv_full %>% count(hiv_result , gender) %>%
  ggplot(aes(hiv_result, n, fill = gender)) +
  geom_col()

```

```{r}
hiv_full %>% count(region)

hiv_full %>% group_by(region) %>% 
  count(hiv_result) %>%
  reframe(region = region,
          result= hiv_result,
          n = n,
          prop = n/(sum(n))) %>%
  ggplot(aes(region, prop, fill = result)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format()) +
  facet_wrap(~ result)

```

# Prepare data for modeling 
## select variables for the model

```{r}
hiv_df <- hiv_full %>% select(hiv_result, region, gender, age, work12mo, religion, schlhi, schlat, occuptn, part12monum, evermar, curmar,	firstsxagedk,	hivtstever,	hivlikely, avoidpreg) %>% as_tibble()

hiv_df <- hiv_df %>% mutate(occupation = case_when(work12mo == 2 ~ "no work",
                                         occuptn == 1 ~ "PROFESSIONAL/TECHNICAL/MANAGERIAL",
                                         occuptn == 2 ~ "CLERICAL",
                                         occuptn == 3 ~ "SALES AND SERVICES",
                                         occuptn == 4 ~ "SKILLED MANUAL",
                                         occuptn == 5 ~ "UNSKILLED MANUAL",
                                         occuptn == 6 ~ "AGRICULTURE",
                                         occuptn == 7 ~ "HOUSE WIFE",
                                         occuptn == 96 ~ "OTHER",
                                         TRUE ~ NA),
                            education = case_when(schlat == 2 ~ "no formal education",
                                                   schlhi == 1 ~ "PRIMARY",
                                                   schlhi == 2 ~ "SECONDARY",
                                                   schlhi == 3 ~ "TECHNICAL/VOCATIONAL",
                                                   schlhi == 4 ~ "HIGHER",
                                                   TRUE ~ NA),
                            sex_part = case_when(firstsxagedk == 96 ~ 0,
                                                 part12monum == 0 ~ 0,
                                                 part12monum == 1 ~ 1,
                                                 part12monum == 2 ~ 2,
                                                 part12monum == 3 ~ 3,
                                                 part12monum == 4 ~ 4, 
                                                 TRUE ~ NA),
                            avoidpreg = case_when(avoidpreg == 2 ~ "No",
                                                  avoidpreg == 1 ~ "Yes",
                                                  TRUE ~ NA),
                            religion = case_when(religion == 1 ~ "ORTHODOX",
                                                 religion == 2 ~ "MUSLIM",
                                                 religion == 4 ~ "PROTESTANT",
                                                 religion == 96|3 ~ "OTHER",
                                                 TRUE ~ NA),
                            marital = case_when(evermar == 2 ~ "NEVER MARRIED",
                                                curmar == 1 ~ "MARRIED",
                                                curmar == 2 ~ "LIVING TOGETHER",
                                                curmar == 3 ~ "WIDOWED",
                                                curmar == 4 ~ "DIVORCED",
                                                curmar == 5 ~ "SEPARATED",
                                                TRUE ~ NA),
                            # evermar = case_when(evermar == 1 ~ "Yes",
                                                #evermar == 2 ~ "No",
                                                #TRUE ~ NA),
                           hivtstever = case_when(hivtstever == 1 ~ "Yes",
                                                   hivtstever == 2 ~ "No",
                                                   TRUE ~ NA),
                            hivlikely = case_when(hivlikely == 1 ~ "VERY LIKELY",
                                                  hivlikely == 2 ~ "SOMEWHAT LIKELY",
                                                  hivlikely == 3 ~ "NOT LIKELY",
                                                  hivlikely == -8 ~ "DON'T KNOW",
                                                  TRUE ~ NA)) %>% mutate_if(is.character, factor) %>%
  select(-c(work12mo,occuptn, schlat, schlhi, part12monum,firstsxagedk, evermar, curmar)) 


#chisq.test(hiv_df$evermar, hiv_df$hiv_result) # X-squared = 177.61, df = 1, p-value < 2.2e-16
#chisq.test(hiv_df$marital, hiv_df$hiv_result) #X-squared = 585.91, df = 5, p-value < 2.2e-16

```


# last EDA
```{r}

```

```{r}
hiv_df %>% select(hiv_result, age, marital, sex_part, education) %>% na.omit() %>% 
  ggpairs(columns = 2:5, aes(color = hiv_result, alpha = 0.3))
```




# Build a Model
## Split the data 

```{r}
set.seed(123)

hiv_split <- initial_split(hiv_df, strata = hiv_result)
hiv_train <- training(hiv_split)
  
hiv_test <- testing(hiv_split)


set.seed(234)

hiv_vfold <- vfold_cv(hiv_train, strata = hiv_result)

metrics_set <- metric_set(roc_auc, accuracy, sens, spec)
```


# Logistic regression
```{r}
glmnet_recipe <- 
  recipe(formula = hiv_result ~ ., data = hiv_train) %>% 
  step_impute_knn(hivtstever, hivlikely, avoidpreg, occupation, education, sex_part, marital) %>%
  step_zv(all_predictors()) %>% 
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors(), -hiv_result) %>%
  step_smote(hiv_result)

glmnet_spec <- 
  logistic_reg(penalty = tune(), mixture = tune()) %>% 
  set_mode("classification") %>% 
  set_engine("glmnet") 

glmnet_workflow <- 
  workflow() %>% 
  add_recipe(glmnet_recipe) %>% 
  add_model(glmnet_spec) 

glmnet_grid <- tidyr::crossing(penalty = 10^seq(-6, -1, length.out = 20), mixture = c(0.05,0.2, 0.4, 0.6, 0.8, 1)) 

glmnet_tune <- 
  tune_grid(glmnet_workflow, 
            resamples = hiv_vfold,
            grid = glmnet_grid,
            metrics = metrics_set) 
```


## Explor the results

```{r}
# collect_metrics(glmnet_tune)
show_best(glmnet_tune,metric =  "roc_auc")
show_best(glmnet_tune, metric =  "accuracy")
show_best(glmnet_tune, metric =  "sens")
show_best(glmnet_tune, metric =  "spec")
```

```{r}
autoplot(glmnet_tune)
```


## finalize the workflow with the best performing parameters

```{r}
final_glm <- glmnet_workflow %>%
  finalize_workflow(select_best(glmnet_tune))

final_glm
```


## fits the finalized model to the training data and evaluates on the testing data

```{r}
hiv_fit_glm <- last_fit(final_glm, hiv_split)
hiv_fit_glm
```


```{r}
collect_metrics(hiv_fit_glm)
```

```{r}
collect_predictions(hiv_fit_glm) %>%
  conf_mat(hiv_result, .pred_class) %>%
  summary()
```


```{r}
collect_predictions(hiv_fit_glm) %>%
  conf_mat(hiv_result, .pred_class) %>%
  autoplot(fill = "midnightblue")
```


## Feature importance

```{r}
imp_spec_glm <- glmnet_spec %>%
  finalize_model(select_best(glmnet_tune)) %>%
  set_engine("glmnet", importance = "permutation")

workflow() %>%
  add_recipe(glmnet_recipe) %>%
  add_model(imp_spec_glm) %>%
  fit(hiv_train) %>%
  pull_workflow_fit() %>%
  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue"))
```



# Decision tree


```{r}
C50_recipe <- 
  recipe(formula = hiv_result ~ ., data = hiv_train)%>%
  step_impute_knn(hivtstever, hivlikely, avoidpreg, occupation, education, sex_part, marital) %>%
  step_zv(all_predictors()) %>% 
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors(), -hiv_result) %>%
  step_smote(hiv_result)

C50_spec <- 
  boost_tree(trees = tune(), min_n = tune()) %>% 
  set_mode("classification") %>% 
  set_engine("C5.0") 

C50_workflow <- 
  workflow() %>% 
  add_recipe(C50_recipe) %>% 
  add_model(C50_spec) 

set.seed(71885)
C50_tune <-
  tune_grid(C50_workflow,
            resamples = hiv_vfold,
            grid = 10,
            metrics = metrics_set)
```


## Explor the results

```{r}
# collect_metrics(C50_tune)
show_best(C50_tune, metric =  "roc_auc")
show_best(C50_tune, metric =  "accuracy")
show_best(C50_tune, metric =  "sens")
show_best(C50_tune, metric =  "spec")
```

```{r}
autoplot(C50_tune)
```


## finalize the workflow with the best performing parameters

```{r}
final_C50 <- C50_workflow %>%
  finalize_workflow(select_best(C50_tune))

final_C50
```


## fits the finalized model to the training data and evaluates on the testing data

```{r}
hiv_fit_C50 <- last_fit(final_C50, hiv_split)
hiv_fit_C50
```


```{r}
collect_metrics(hiv_fit_C50)
```

```{r}
collect_predictions(hiv_fit_C50) %>%
  conf_mat(hiv_result, .pred_class) %>%
  summary()
```

```{r}
collect_predictions(hiv_fit_C50) %>%
  conf_mat(hiv_result, .pred_class) %>%
  autoplot(fill = "midnightblue")
```

#### Feature importance 

```{r}
imp_spec_C50 <- C50_spec %>%
  finalize_model(select_best(C50_tune)) %>%
  set_engine("C5.0", importance = "permutation")

workflow() %>%
  add_recipe(C50_recipe) %>%
  add_model(imp_spec_C50) %>%
  fit(hiv_train) %>%
  pull_workflow_fit() %>%
  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue"))
```



# Random forest

```{r}
ranger_recipe <- 
  recipe(formula = hiv_result ~ ., data = hiv_train) %>%
  step_impute_knn(hivtstever, hivlikely, avoidpreg, occupation, education, sex_part, marital) %>%
  step_zv(all_predictors()) %>% 
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors(), -hiv_result) %>%
  step_smote(hiv_result)

ranger_spec <- 
  rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>% 
  set_mode("classification") %>% 
  set_engine("ranger") 

ranger_workflow <- 
  workflow() %>% 
  add_recipe(ranger_recipe) %>% 
  add_model(ranger_spec) 

set.seed(28898)
ranger_tune <-
  tune_grid(ranger_workflow,
            resamples = hiv_vfold,
            grid = 10,
            metrics = metrics_set)
```


## Explor the results

```{r}
# collect_metrics(ranger_tune)
show_best(ranger_tune, metric =  "roc_auc")
show_best(ranger_tune, metric =  "accuracy")
show_best(ranger_tune, metric =  "sens")
show_best(ranger_tune, metric =  "spec")
```

```{r}
autoplot(ranger_tune)
```

## finalize the workflow with the best performing parameters

```{r}
final_rf <- ranger_workflow %>%
  finalize_workflow(select_best(ranger_tune))

final_rf
```



## fits the finalized model to the training data and evaluates on the testing data

```{r}
hiv_fit_rf <- last_fit(final_rf, hiv_split)
hiv_fit_rf
```


```{r}
collect_metrics(hiv_fit_rf)
```


```{r}
collect_predictions(hiv_fit_rf) %>%
  conf_mat(hiv_result, .pred_class) %>%
  summary()
```

```{r}
collect_predictions(hiv_fit_rf) %>%
  conf_mat(hiv_result, .pred_class) %>%
  autoplot(fill = "midnightblue")
```


## Feature importance

```{r}
imp_spec_rf <- ranger_spec %>%
  finalize_model(select_best(ranger_tune)) %>%
  set_engine("ranger", importance = "permutation")

workflow() %>%
  add_recipe(ranger_recipe) %>%
  add_model(imp_spec_rf) %>%
  fit(hiv_train) %>%
  pull_workflow_fit() %>%
  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue"))
```




## Xgboost

```{r}
xgboost_recipe <- 
  recipe(formula = hiv_result ~ ., data = hiv_train) %>% 
  step_impute_knn(hivtstever, hivlikely, avoidpreg, occupation, education, sex_part, marital) %>%
  step_zv(all_predictors()) %>% 
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors(), -hiv_result) %>%
  step_smote(hiv_result) 

xgboost_spec <- 
  boost_tree(trees = tune(), min_n = tune(), tree_depth = tune(), learn_rate = tune(), 
    loss_reduction = tune(), sample_size = tune()) %>% 
  set_mode("classification") %>% 
  set_engine("xgboost") 

xgboost_workflow <- 
  workflow() %>% 
  add_recipe(xgboost_recipe) %>% 
  add_model(xgboost_spec) 

set.seed(25533)
xgboost_tune <-
  tune_grid(xgboost_workflow, 
            resamples = hiv_vfold, 
            grid = 10,
            metrics = metrics_set)
```


```{r}
show_best(xgboost_tune, metric =  "roc_auc")
show_best(xgboost_tune, metric =  "accuracy")
```



## finalize the workflow with the best performing parameters

```{r}
final_xg <- xgboost_workflow %>%
  finalize_workflow(select_best(xgboost_tune))

final_xg
```



## fits the finalized model to the training data and evaluates on the testing data

```{r}
hiv_fit_xg <- last_fit(final_xg, hiv_split)
hiv_fit_xg
```


```{r}
collect_metrics(hiv_fit_xg)
```


```{r}
collect_predictions(hiv_fit_xg) %>%
  conf_mat(hiv_result, .pred_class) %>%
  summary()
```

```{r}
collect_predictions(hiv_fit_xg) %>%
  conf_mat(hiv_result, .pred_class) %>%
  autoplot(fill = "midnightblue")
```


## Feature importance

```{r}
imp_spec_xg <- xgboost_spec %>%
  finalize_model(select_best(xgboost_tune)) %>%
  set_engine("xgboost", importance = "permutation")

workflow() %>%
  add_recipe(xgboost_recipe) %>%
  add_model(imp_spec_xg) %>%
  fit(hiv_train) %>%
  pull_workflow_fit() %>%
  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue"))
```


